---
title: "Analyzing Changes in Soil Organic Matter During Elevated CO₂ Experiments"
author: "Ananda Kurth"
date: "2025-03-03"
output: 
  html_document:
    toc: true
    toc_float: true
---

# Chapter 4 - Report Exercises

## 1. Introduction

This exercise compares Soil organic carbon (SOC) measurements from carbon dioxide (CO₂) enrichment experiments.

## 2. Library Loading

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(tidyr)
library(readr)
library(stringr)
library(purrr)
library(openxlsx)
library(zoo)
library(ggplot2)
```

## 3. Raw Data Import

```{r load-dataset}
raw_data <- read.xlsx(here::here("data/1249534s1-s6.xlsx"))
```

## 4. Data Cleaning

```{r clean-data}
cleaned_data <- as_tibble(raw_data) |>
  rename(
    experiment           = Database.S1..Overview.of.CO2.enrichment.studies.reporting.soil.C.contents.that.were.used.in.our.analysis.,
    citation             = X2,
    depth                = X3,
    sample_date          = X4,
    time_years           = X5,
    SOC_ambient          = X6,
    SOC_elevated         = X7,
    n_sample_SOC_ambient = X8,
    n_sample_SOC_elevated = X9,
    description_source   = X10,
    value_treatment      = X11
  ) |>                    # Rename columns for clarity
  slice(-c(1:3), ) |>    # Remove top rows that contain header metadata
  select(experiment,
         time_years,
         starts_with("SOC"),
         starts_with("n_sample")) |>                    # Keep only the columns needed for our analysis
  mutate(
    time_years  = as.numeric(time_years),
    SOC_ambient = as.numeric(SOC_ambient),
    SOC_elevated = as.numeric(SOC_elevated)
  ) |>                    # Convert these to numeric
  mutate(experiment = na.locf(na.locf(experiment, na.rm = FALSE), fromLast = TRUE)) |>                    # Fill missing experiment names downward
  filter(!is.na(experiment))  # Drop any stray rows without an experiment name
```

## 5. Save Cleaned Data to CSV

```{r save-dataset}
write_csv(cleaned_data, file = here::here("data/cleaned_data.csv"))
head(cleaned_data)
```

## 6. Compute Log-Response Ratio & Define Phases

```{r log-response}
SOC_data <- read_csv(here::here("data/cleaned_data.csv"), show_col_types = FALSE) |>
  mutate(log_response_ratio = log(SOC_elevated / SOC_ambient)) |>                     # Compute LRR for each row
  mutate(
    phase = case_when(
      time_years < 3                ~ "early_phase",
      time_years >= 3 & time_years <= 6 ~ "mid_phase",
      time_years > 6                ~ "late_phase"
    )
  )                       # Assign a phase label
```

## 7. Phase-Wise Aggregation

Within each phase, I compute the mean log-response ratio to see the average effect across all experiments.

```{r mean-log-response}
phase_aggregate <- SOC_data |>
  group_by(phase) |>
  summarize(mean_log_response_ratio = mean(log_response_ratio, na.rm = TRUE))

phase_aggregate |>
  knitr::kable()
```

## 8. Visualize Mean LRR by Phase

```{r visualize-results}
ggplot(phase_aggregate,
       aes(x = phase, y = mean_log_response_ratio, fill = phase)) +
  geom_bar(stat = "identity") +
  labs(title = "Mean Log-Response Ratio by Experimental Phase", x     = "Experimental Phase", y     = "Mean Log-Response Ratio") +
  scale_x_discrete(limits = c("early_phase", "mid_phase", "late_phase")) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title      = element_text(hjust = 0.5))
```

## 9. Interpretation & Discussion

### What are the data that you are looking at?

We have Soil organic carbon (SOC) measurements from multiple carbon dioxide (CO₂) enrichment experiments. Each record includes SOC under ambient and elevated CO₂, and the duration (in years) since experiment start.

### What do you expect your analysis to show, what is your hypothesis? How should soil organic matter content change under elevated CO2?

Elevated levels of CO₂ generally increases plant growth, which in turn increases the amount of carbon being deposited into the soil. As a result, I expect that the SOC in conditions of elevated CO₂ will surpass that of ambient conditions.

### Interpret your results after aggregating the data: What do your final numbers mean?

-   Early Phase (t \< 3 years): mean LRR ≈ +0.0125 → a \~1.25% average increase in SOC under elevated CO₂.

-   Mid Phase (3 ≤ t ≤ 6 years): mean LRR ≈ +0.0173 → a \~1.73% average increase, slightly larger.

-   Late Phase (t \> 6 years): mean LRR ≈ –0.0306 → a \~3.06% decrease under elevated CO₂.

In the early phases, SOC levels are higher under elevated CO₂ conditions, but this reverses in the late phase, causing SOC to fall below ambient levels.

### Do they support your initial hypothesis? Why so, why not?Does this support our hypothesis?

The results partially align with my hypothesis. The early and mid phases do show an increase in SOC, consistent with my expectation that elevated CO₂ boosts soil carbon through higher plant inputs. However, the late phase shows a negative LRR. This indicates that any initial gain in SOC may be reversed after about 6 years, likely because microbial decomposition and carbon turnover accelerate under elevated CO₂ (van Groenigen et al. (2014)).
