---
title: "6.4_exercises"
output: html_document
---

# Exercises chapter 6
```{r libraries}
library(dplyr)
library(jsonlite)
library(ggplot2)
library(readr)
library(lubridate)
```

## Files and file formats 

### Reading and writing human-readable files

1.  While **not leaving your R session**, download and open the files at the following locations: <https://raw.githubusercontent.com/geco-bern/agds_book/refs/heads/main/book/data/demo_1.csv> <https://raw.githubusercontent.com/geco-bern/agds_book/refs/heads/main/book/data/demo_2.csv> <https://raw.githubusercontent.com/geco-bern/agds_book/refs/heads/main/book/data/demo_3.csv>

```{r download}

# read in the first demo
demo_01 <- read.table(
    "https://raw.githubusercontent.com/geco-bern/agds_book/refs/heads/main/book/data/demo_1.csv",
    sep = ",",
    header = TRUE
    )

# read in the second demo
demo_02 <- read.table(
    "https://raw.githubusercontent.com/geco-bern/agds_book/refs/heads/main/book/data/demo_2.csv",
    sep = ",",
    header = TRUE
    )

# read in the third demo
demo_03 <- read.table(
    "https://raw.githubusercontent.com/geco-bern/agds_book/refs/heads/main/book/data/demo_3.csv",
    sep = ",",
    header = TRUE
)
```

2.  Once loaded into your R environment, combine and save all data as a temporary CSV file. Read in the new temporary CSV file, and save it as a JSON file in your current working directory.

```{r combine}
demo_all <- bind_rows(demo_01, demo_02, demo_03)

# writing the data to a temporary CSV file
write.table(
    demo_all, 
    file = file.path(tempdir(), "tmp_csv_file.csv"),
    col.names = TRUE,
    row.names = FALSE,
    sep = ","
)

# reading the data from the temporary CSV file
demo_all_reload <- read.table(
    file.path(tempdir(), "tmp_csv_file.csv"),
    sep = ",",
    header = TRUE
)

# writing the data to a JSON file and save in the current working directory
write_json(
    demo_all_reload,
    path = file.path(tempdir(), "tmp_json_file.json")
)  

df_json <- read_json(
  file.path(tempdir(), "tmp_json_file.json"),
  simplifyVector = TRUE
)

identical(df_json, demo_all_reload)
```

## Reading and writing binary files

Download and open the following file: <https://raw.githubusercontent.com/geco-bern/agds_book/refs/heads/main/book/data/demo_data.nc>

1.  What file format are we dealing with? \> NetCDF - Network Common Data Form. \> NetCDF is a file format for storing multidimensional scientific data (variables) such as temperature, humidity, pressure, wind speed, and direction.\
2.  What library would you use to read this kind of data? \> {ncdf4}
3.  What does this file contain?

```{r}
library(ncdf4)
nc <- nc_open("https://raw.githubusercontent.com/geco-bern/agds_book/refs/heads/main/book/data/demo_data.nc")
nc
```

4.  Write this file to disk in a different geospatial format you desire (use the R documentation of the library used to read the file and the chapter information).

```{r write}
# write the data as a geotiff
terra::writeRaster(
  nc,
  filename = "./test.tif",
  overwrite = TRUE
  ) 
```

5.  Download and open the following file: <https://raw.githubusercontent.com/geco-bern/agds_book/refs/heads/main/book/data/demo_data.tif>.

```{r}
library("terra")
url <- "https://raw.githubusercontent.com/geco-bern/agds_book/refs/heads/main/book/data/demo_data.tif"

demo_data <- terra::rast(url)
demo_data
```

6.  Does this data seem familiar, and how can you tell? What are your conclusions?

## API use

### GET

1.  Download the HWSD total sand content data for the extent of Switzerland following the tutorial example.

```{r}
# set API URL endpoint
# for the total sand content
url <- "https://thredds.daac.ornl.gov/thredds/ncss/ornldaac/1247/T_SAND.nc4"

# formulate query to pass to httr
query <- list(
  "var" = "T_SAND",
  "south" = 32,
  "west" = -81,
  "east" = -80,
  "north" = 34,
  "disableProjSubset" = "on",
  "horizStride" = 1,
  "accept" = "netcdf4"
)

# download data using the
# API endpoint and query data
status <- httr::GET(
  url = url,
  query = query,
  httr::write_disk(
    path = file.path(tempdir(), "T_SAND.nc"),
    overwrite = TRUE
  )
)

```

Visualize/plot the data as a simple map.

```{r}
sand <- terra::rast(file.path(tempdir(), "T_SAND.nc"))
terra::plot(sand)
```

Download the HWSD topsoil silt content for the extent of Switzerland.

```{r}
# set API URL endpoint
# for the topsoil silt content
url <- "https://thredds.daac.ornl.gov/thredds/ncss/ornldaac/1247/T_SILT.nc4"

# formulate query to pass to httr
query <- list(
  "var" = "T_SILT",
  "south" = 32,
  "west" = -81,
  "east" = -80,
  "north" = 34,
  "disableProjSubset" = "on",
  "horizStride" = 1,
  "accept" = "netcdf4"      
)

status <- httr::GET(
  url = url,
  query = query,
  httr::write_disk(
    path = file.path(tempdir(), "T_SILT.nc"),
    overwrite = TRUE
  )
)

silt <- terra::rast(file.path(tempdir(), "T_SILT.nc"))
terra::plot(silt)
```

### 6.4.0.1 Dedicated library

1.  Use the {hwsdr} library (a dedicated package for the API) to download the same data. How does this compare to the previous code written?
2.  List how many data products there are on the ORNL MODIS data repository.
3.  Download the MODIS land cover map for the canton of Bern.

```{r}
library("hwsdr")
library("MODISTools")

```
